Topological Equivariant Artist Model for Visualization Library Architecture

### 2022-07-28 paper/main.tex
intro

\IEEEraisesectionheading{\section{Introduction}\label{sec:intro}}
\note{need a math check + rough cut on sections 4+}

\note{Spelling out vis library component behavior w/ alg top + cat theory lets us be super general in terms of inputs/outputs (what data, what vis) while super specific about the constraints the components need to satisfy to transform that input to that output in a structure preseving way}

\IEEEPARstart{V}isualizations are expected to be a faithful translation of the inherent structure of the data being visualized; therefore the components of visualization software libraries that translate data to graphics are also expected to preserve this structure. Motivated by a need to develop new library components that adapt to modern data needs, especially high dimensional multivariate and distributed data, we propose a framework that topological and field structure of the input data that must be mapped to the visualization and which changes to data must be expressed in visualization, and a a functional design framework that describes composability in terms of functional algebraic operations. \note{this may be a good place to incorporate types} This allows complex visualizations to be built from simple verifiable parts\cite{huHowFunctionalProgramming2015, hughesWhyFunctionalProgramming1989}, which makes these components easier to drop into the existing library and be used by domain specific library developers to build new libraries that work with existing ones.  While many of these elements have been partially described elsewhere, our framework naturally combines them in a unified framework based on ideas in category theory. Further we see the application of these ideas in low level visualization library design as our motivation, and we hope these ideas will help shape the reorganization of critical data visualization libraries, such as Matplotlib. The contribution of this paper is a uniform method for describing the structure in data and a mathematical framework for describing, constructing, and testing structure preserving visualization library components.

>>>>

\IEEEraisesectionheading{\section{Introduction}\label{sec:intro}}

\IEEEPARstart{V}isualizations are expected to be a faithful translation of data, meaning that inherent structure in the data should be present in the visualization; therefore the components of visualization software libraries that translate data to graphics are also expected to be structure preserving. This inherent structure can be quite complex, such as when the data is high dimensional, multivariate, deeply nested, or encoded in a distributed manner. We propose that we can rigorously specify what structure the library components are expected to preserve using category theory. We also propose that these specifications can generalize to most structure types by extending previous work on encoding  inherent structure in a consistent manner using mathematical structures from algebraic topology. These algebraic structures allow us to encode inherent structure as programmatic types, and the use of category theory facilitates developing a functional design framework that describes composability in terms of functional algebraic operations. The typing system and composability inherent in a functional design encourage library developers to build complex visualization components from simple verifiable parts that have few side effects by design\cite{huHowFunctionalProgramming2015, hughesWhyFunctionalProgramming1989}. These categorically specified components could be built as a standalone library, but can also be integrated into existing libraries because they are inherently self-contained. Furthermore we see the application of these ideas in low level visualization library design as our motivation, and we hope these ideas will help shape the reorganization of critical data visualization libraries, such as Matplotlib. The contribution of this paper is a mathematical framework for describing structure preserving visualization components in a manner that is generalizable to different types of inherent structure. This framework provides also provides guidence for the construction and testing of structure preserving visualization library components.
